---
title: "AFM 423 Final Project"
author: "Xuntian Lin 20763179"
output: pdf_document
date: "2023-04-12"
---


Import the required dataset
```{r}
# install.packages("readxl")
# Load the library
library("readxl")

# Load the data and modify the column names
data_ml <- read_excel("data_ml.xlsx")
# load("data_ml.RData")
colnames(data_ml)[2] <- "date"
```


```{r}

# install.packages("glmnet")
library(glmnet)
library(dplyr)
library(tidyverse)
library(lubridate)

data_ml <- data_ml %>% 
    filter(date > "1999-12-31",         # Keep the date with sufficient data points
           date < "2019-01-01") %>%
    mutate(LMH = Mkt_Cap_3M_Usd / Share_Turn_3M) %>%
    arrange(stock_id, date)             # Order the data

# List of features of data_ml (i.e. column names)
features <- colnames(data_ml[3:95])   # Keep the feature's column names

stock_ids <- levels(as.factor(data_ml$stock_id)) # A list of all stock_ids
stock_days <- data_ml %>%                        # Compute the number of data points per stock
    group_by(stock_id) %>% summarize(nb = n()) 
stock_ids_short <- stock_ids[which(stock_days$nb == max(stock_days$nb))] # Stocks with full data

returns <- data_ml %>%                           # Compute returns, in matrix format, in 3 steps:
    filter(stock_id %in% stock_ids) %>%    # 1. Filtering the data
    dplyr::select(date, stock_id, R1M_Usd) %>%   # 2. Keep returns along with dates & firm names
    pivot_wider(names_from = "stock_id", 
                values_from = "R1M_Usd")         # 3. Put in matrix shape 
```


Load the Fama French 3 factor model dataset
```{r}
library(quantmod)
library(xtable)

min_date <- "1963-07-31"
max_date <- "2020-03-28"

ff_factors <- read_csv("F-F_Research_Data_Factors.CSV", skip = 3) %>%
  rename(date = 1, MKT_RF = 'Mkt-RF') %>%
  mutate_at(vars(-date), as.numeric) %>%
  mutate(date = ymd(parse_date_time(date, "%Y%m"))) %>%
  mutate(date = rollback(date + months(1))) #%>%
  # mutate(monthid = (year(date) - 2000) * 12 + month(date))


ff_factors <- ff_factors %>%
  mutate(MKT_RF = MKT_RF / 100, # Scale returns
         SMB = SMB / 100,
         HML = HML / 100,
         RF = RF/100) %>%
  filter(date >= min_date, date <= max_date)    # Finally, keep only recent points

knitr::kable(head(ff_factors), booktabs = TRUE,
caption = "Sample of monthly factor returns.") # A look at the data (see table)

```


Visualization of the ff_factors
```{r}
ff_factors %>%
  mutate(date = year(date)) %>%
  gather(key = factor, value = value, -date) %>%
  group_by(date, factor) %>%
  summarise(value = mean(value)) %>%
  ggplot(aes(x = date, y = value, color = factor)) + 
  geom_line() + coord_fixed(500) + theme_light()
```


Create extra monthid columns
```{r}
#ff_3$monthid <- (year(ff_3$dateff) - 2000) * 12 + month(ff_3$dateff)

ml_returns <-
  data_ml %>%
  select(date, stock_id, R1M_Usd) %>%
  # mutate(monthid = (year(date) - 2000) * 12 + month(date)) %>%
  left_join(ff_factors, by = "date") %>%
  mutate(R_excess = R1M_Usd - RF) # %>%
  # select( -monthid)         # Drop the unnecessary columns here
```


# CAPM Model
```{r}
nb_factors <- 1                                                     # Number of factors
data_FM <- left_join(data_ml %>%                                    # Join the 2 datasets
                         dplyr::select(date, stock_id, R1M_Usd), # (with returns...
                         # filter(stock_id %in% stock_ids_short),     # ... over some stocks)
                     ff_factors, 
                     by = "date") %>% 
    group_by(stock_id) %>%                                          # Grouping
    mutate(R1M_Usd = lag(R1M_Usd)) %>%                              # Lag returns
    ungroup() %>%
    na.omit() %>%                                                   # Remove missing points
    pivot_wider(names_from = "stock_id", values_from = "R1M_Usd")

models <- lapply(paste0("`", stock_ids, 
                        '` ~  MKT_RF'),           # Model spec
                 function(f){ lm(as.formula(f), data = data_FM,           # Call lm(.)
                                 na.action="na.exclude") %>%       
                         summary() %>%                                    # Gather the output
                         "$"(coef) %>%                                    # Keep only coefs
                         data.frame() %>%                                 # Convert to dataframe
                         dplyr::select(Estimate)}                         # Keep the estimates
                 )
betas <- matrix(unlist(models), ncol = nb_factors + 1, byrow = T) %>%     # Extract the betas
    data.frame(row.names = stock_ids)                               # Format: row names
colnames(betas) <- c("Constant", "MKT_RF")    # Format: col names
```

```{r}
loadings <- betas %>%                            # Start from loadings (betas)
    dplyr::select(-Constant) %>%                 # Remove constant
    data.frame()                                 # Convert to dataframe             
ret <- returns %>%                               # Start from returns
    dplyr::select(-date) %>%                     # Keep the returns only
    data.frame(row.names = returns$date) %>%     # Set row names
    t()                                          # Transpose
FM_data <- cbind(loadings, ret)                  # Aggregate both
```

Then we do the second round of regression
```{r}
models <- lapply(paste("`", returns$date, "`", ' ~  MKT_RF', sep = ""),
function(f){ lm(as.formula(f), data = FM_data) %>%                        # Call lm(.)
                         summary() %>%                                    # Gather the output
                         "$"(coef) %>%                                    # Keep only the coefs
                         data.frame() %>%                                 # Convert to dataframe
                         dplyr::select(Estimate)}                         # Keep only estimates
                 )
gammas <- matrix(unlist(models), ncol = nb_factors + 1, byrow = T) %>%    # Switch to dataframe
    data.frame(row.names = returns$date)                                  # & set row names
colnames(gammas) <- c("Constant", "MKT_RF")   # Set col names
```

Visually the gamma
```{r}
gammas[2:nrow(gammas),] %>%                                         # Take gammas:
    # The first row is omitted because the first row of returns is undefined
    dplyr::select(MKT_RF) %>%                             # Select 3 factors
    bind_cols(date = data_FM$date) %>%                              # Add date
    gather(key = factor, value = gamma, -date) %>%                  # Put in tidy shape
    ggplot(aes(x = date, y = gamma, color = factor)) +              # Plot
    geom_line() + facet_grid( factor~. ) +                          # Lines & facets
    scale_color_manual(values=c("#F87E1F", "#0570EA", "#F81F40")) + ylim(-0.5, 0.5) # + # Colors
    # coord_fixed() + theme_light()                                # Fix x/y ratio
```


# Simple Fama French three factor model
```{r}
nb_factors <- 3                                                     # Number of factors
data_FM <- left_join(data_ml %>%                                    # Join the 2 datasets
                         dplyr::select(date, stock_id, R1M_Usd), # (with returns...
                         # filter(stock_id %in% stock_ids_short),     # ... over some stocks)
                     ff_factors, 
                     by = "date") %>%
    group_by(stock_id) %>%                                           # Grouping
    mutate(R1M_Usd = lag(R1M_Usd)) %>%                              # Lag returns
    ungroup() %>%
    na.omit() %>%                                                   # Remove missing points
    pivot_wider(names_from = "stock_id", values_from = "R1M_Usd")

models <- lapply(paste0("`", stock_ids, 
                        '` ~  MKT_RF + SMB + HML'),           # Model spec
                 function(f){ lm(as.formula(f), data = data_FM,           # Call lm(.)
                                 na.action="na.exclude") %>%       
                         summary() %>%                                    # Gather the output
                         "$"(coef) %>%                                    # Keep only coefs
                         data.frame() %>%                                 # Convert to dataframe
                         dplyr::select(Estimate)}                         # Keep the estimates
                 )
betas <- matrix(unlist(models), ncol = nb_factors + 1, byrow = T) %>%     # Extract the betas
    data.frame(row.names = stock_ids)                               # Format: row names
colnames(betas) <- c("Constant", "MKT_RF", "SMB", "HML")    # Format: col names
```


```{r}
loadings <- betas %>%                            # Start from loadings (betas)
    dplyr::select(-Constant) %>%                 # Remove constant
    data.frame()                                 # Convert to dataframe             
ret <- returns %>%                               # Start from returns
    dplyr::select(-date) %>%                     # Keep the returns only
    data.frame(row.names = returns$date) %>%     # Set row names
    t()                                          # Transpose
FM_data <- cbind(loadings, ret)                  # Aggregate both
```


Then we do the second round of regression
```{r}
models <- lapply(paste("`", returns$date, "`", ' ~  MKT_RF + SMB + HML', sep = ""),
function(f){ lm(as.formula(f), data = FM_data) %>%                        # Call lm(.)
                         summary() %>%                                    # Gather the output
                         "$"(coef) %>%                                    # Keep only the coefs
                         data.frame() %>%                                 # Convert to dataframe
                         dplyr::select(Estimate)}                         # Keep only estimates
                 )

gammas <- matrix(unlist(models), ncol = nb_factors + 1, byrow = T) %>%    # Switch to dataframe
    data.frame(row.names = returns$date)                                  # & set row names
colnames(gammas) <- c("Constant", "MKT_RF", "SMB", "HML")   # Set col names
```

Visually plot the gamma
```{r}
gammas[2:nrow(gammas),] %>%                                         # Take gammas:
    # The first row is omitted because the first row of returns is undefined
    dplyr::select(MKT_RF, SMB, HML) %>%                             # Select 3 factors
    bind_cols(date = data_FM$date) %>%                              # Add date
    gather(key = factor, value = gamma, -date) %>%                  # Put in tidy shape
    ggplot(aes(x = date, y = gamma, color = factor)) +              # Plot
    geom_line() + facet_grid( factor~. ) +                          # Lines & facets
    scale_color_manual(values=c("#F87E1F", "#0570EA", "#F81F40")) #+ # Colors
    #coord_fixed(980) + theme_light()                                # Fix x/y ratio
```


Calculate the statistics
```{r}

library(knitr)
# Gather the main statistics
stats <-
  gammas %>%
  summarise(mean(MKT_RF), mean(SMB), mean(HML), sd(MKT_RF), sd(SMB), sd(HML)) %>%
  round(3)

stats

# Calculate t-stats
r <- t.test(gammas$MKT_RF)
r$statistic
r <- t.test(gammas$SMB)
r$statistic
r <- t.test(gammas$HML)
r$statistic

# Calculate coefficients
cor(gammas$MKT_RF, gammas$SMB)
cor(gammas$MKT_RF, gammas$HML)
cor(gammas$SMB, gammas$HML)
```

# Introduce a new factor called LMH based on share_turn_3m
```{r}
# First put the portfolio in three ones, L, M and H based on the shares
ml_returns_siliq <-
  data_ml %>%
  select(c("date", "stock_id", "R1M_Usd", "Pb", "Mkt_Cap_3M_Usd", "Share_Turn_3M", "Share_Turn_6M", "Share_Turn_12M"))  %>%
  group_by(date) %>%
  mutate(Turn_rank = ntile(Share_Turn_12M, 3)) %>%
  ungroup()

# Take the weighted mean on the cut portfolios
siliq_view <-
  ml_returns_siliq %>%
  group_by(date, Turn_rank) %>%
  summarise(turn_rank_ret = mean(R1M_Usd), .groups = 'drop') %>%
  ungroup()
  

# Create the new factor
seliq_lmh <-
  siliq_view %>%
  group_by(date) %>%
  summarise(LMH = turn_rank_ret[Turn_rank==1] - turn_rank_ret[Turn_rank==3] ) %>%
  mutate_if(is.numeric, round, 4)

siliq_factors <- 
  left_join(seliq_lmh, ff_factors, by = 'date')

# Plot the new variable LMH
siliq_factors %>%
  mutate(date = year(date)) %>%
  gather(key = factor, value = value, -date) %>%
  group_by(date, factor) %>%
  summarise(value = mean(value)) %>%
  ggplot(aes(x = date, y = value, color = factor)) + 
  geom_line()  + theme_light()
```


# Normal model ff-3
```{r}
ml_returns_ff <-
  data_ml %>%
  select(c("date", "stock_id", "R1M_Usd", "Pb", "Mkt_Cap_3M_Usd", "Share_Turn_3M"))  %>%
  mutate(Bp = 1/Pb) %>%
  left_join(ff_factors, by = "date") %>%
  mutate(R_excess = R1M_Usd - RF)
```

```{r}
ff_dplyr_byhand <-
  ml_returns_ff %>%
  lm(R_excess ~ MKT_RF + SMB + HML, 
                data = .) %>% 
  tidy(conf.level = .95, conf.int = T)

ff_dplyr_byhand %>% 
  mutate_if(is.numeric, funs(round(., 3)))

ff_dplyr_byhand %>% 
  mutate_if(is.numeric, funs(round(., 3))) %>%
  filter(term != "(Intercept)") %>% 
  ggplot(aes(x = term, y = estimate, shape = term, color = term)) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high)) +
  labs(title = "FF 3-Factor Coefficients for Our Portfolio",
       subtitle = "nothing in this post is investment advice",
       x = "",
       y = "coefficient",
       caption = "data source: Fama French website") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption  = element_text(hjust = 0))
```


# Four factor model (SiLiq)
```{r}
ml_returns_siliq <-
  data_ml %>%
  select(c("date", "stock_id", "R1M_Usd", "Pb", "Mkt_Cap_3M_Usd", "Share_Turn_3M"))  %>%
  mutate(Bp = 1/Pb) %>%
  left_join(siliq_factors, by = "date") %>%
  mutate(R_excess = R1M_Usd - RF)
```


```{r}
ff_dplyr_byhand <-
  ml_returns_siliq %>%
  lm(R_excess ~ MKT_RF + SMB + LMH, 
                data = .) %>% 
  tidy(conf.level = .95, conf.int = T)

ff_dplyr_byhand %>% 
  mutate_if(is.numeric, funs(round(., 3)))

ff_dplyr_byhand %>% 
  mutate_if(is.numeric, funs(round(., 3))) %>%
  filter(term != "(Intercept)") %>% 
  ggplot(aes(x = term, y = estimate, shape = term, color = term)) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high)) +
  labs(title = "FF 3-Factor Coefficients for Our Portfolio",
       subtitle = "nothing in this post is investment advice",
       x = "",
       y = "coefficient",
       caption = "data source: Fama French website") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption  = element_text(hjust = 0))
```


# (DiLiq)
```{r}
ff_dplyr_byhand <-
  ml_returns_siliq %>%
  lm(R_excess ~ MKT_RF + HML + LMH, 
                data = .) %>% 
  tidy(conf.level = .95, conf.int = T)

ff_dplyr_byhand %>% 
  mutate_if(is.numeric, funs(round(., 3)))

ff_dplyr_byhand %>% 
  mutate_if(is.numeric, funs(round(., 3))) %>%
  filter(term != "(Intercept)") %>% 
  ggplot(aes(x = term, y = estimate, shape = term, color = term)) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high)) +
  labs(title = "FF 3-Factor Coefficients for Our Portfolio",
       subtitle = "nothing in this post is investment advice",
       x = "",
       y = "coefficient",
       caption = "data source: Fama French website") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption  = element_text(hjust = 0))
```

# Siliq, variant 1

```{r}
nb_factors <- 3                                                     # Number of factors
data_FM <- left_join(data_ml %>%                                    # Join the 2 datasets
                         dplyr::select(date, stock_id, R1M_Usd), # (with returns...
                         # filter(stock_id %in% stock_ids_short),     # ... over some stocks)
                     siliq_factors, 
                     by = "date") %>% 
    group_by(stock_id) %>%                                           # Grouping
    mutate(R1M_Usd = lag(R1M_Usd)) %>%                              # Lag returns
    ungroup() %>%
    na.omit() %>%                                                   # Remove missing points
    pivot_wider(names_from = "stock_id", values_from = "R1M_Usd")

models <- lapply(paste0("`", stock_ids, 
                        '` ~  MKT_RF + SMB + LMH'),           # Model spec
                 function(f){ lm(as.formula(f), data = data_FM,           # Call lm(.)
                                 na.action="na.exclude") %>%       
                         summary() %>%                                    # Gather the output
                         "$"(coef) %>%                                    # Keep only coefs
                         data.frame() %>%                                 # Convert to dataframe
                         dplyr::select(Estimate)}                         # Keep the estimates
                 )
betas <- matrix(unlist(models), ncol = nb_factors + 1, byrow = T) %>%     # Extract the betas
    data.frame(row.names = stock_ids)                               # Format: row names
colnames(betas) <- c("Constant", "MKT_RF", "SMB", "LMH")    # Format: col names
```


```{r}
loadings <- betas %>%                            # Start from loadings (betas)
    dplyr::select(-Constant) %>%                 # Remove constant
    data.frame()                                 # Convert to dataframe             
ret <- returns %>%                               # Start from returns
    dplyr::select(-date) %>%                     # Keep the returns only
    data.frame(row.names = returns$date) %>%     # Set row names
    t()                                          # Transpose
FM_data <- cbind(loadings, ret)                  # Aggregate both
```


Then we do the second round of regression
```{r}
models <- lapply(paste("`", returns$date, "`", ' ~  MKT_RF + SMB + LMH', sep = ""),
function(f){ lm(as.formula(f), data = FM_data) %>%                        # Call lm(.)
                         summary() %>%                                    # Gather the output
                         "$"(coef) %>%                                    # Keep only the coefs
                         data.frame() %>%                                 # Convert to dataframe
                         dplyr::select(Estimate)}                         # Keep only estimates
                 )
gammas <- matrix(unlist(models), ncol = nb_factors + 1, byrow = T) %>%    # Switch to dataframe
    data.frame(row.names = returns$date)                                  # & set row names
colnames(gammas) <- c("Constant", "MKT_RF", "SMB", "LMH")   # Set col names
```

Visually the gamma
```{r}
gammas[2:nrow(gammas),] %>%                                         # Take gammas:
    # The first row is omitted because the first row of returns is undefined
    dplyr::select(MKT_RF, SMB, LMH) %>%                             # Select 3 factors
    bind_cols(date = data_FM$date) %>%                              # Add date
    gather(key = factor, value = gamma, -date) %>%                  # Put in tidy shape
    ggplot(aes(x = date, y = gamma, color = factor)) +              # Plot
    geom_line() + facet_grid( factor~. ) +                          # Lines & facets
    scale_color_manual(values=c("#F87E1F", "#0570EA", "#F81F40")) #+ # Colors
    #coord_fixed(980) + theme_light()                                # Fix x/y ratio
```

Calculate the statistics
```{r}
library(knitr)
# Gather the main statistics
stats <-
  gammas %>%
  summarise(mean(MKT_RF), mean(SMB), mean(LMH), sd(MKT_RF), sd(SMB), sd(LMH)) %>%
  round(3)

stats

# Calculate t-stats
r <- t.test(gammas$MKT_RF)
r$statistic
r <- t.test(gammas$SMB)
r$statistic
r <- t.test(gammas$LMH)
r$statistic

# Calculate coefficients
cor(gammas$MKT_RF, gammas$SMB)
cor(gammas$MKT_RF, gammas$LMH)
cor(gammas$SMB, gammas$LMH)
```



# Diliq
```{r}
nb_factors <- 3                                                     # Number of factors
data_FM <- left_join(data_ml %>%                                    # Join the 2 datasets
                         dplyr::select(date, stock_id, R1M_Usd), # (with returns...
                         # filter(stock_id %in% stock_ids_short),     # ... over some stocks)
                     siliq_factors, 
                     by = "date") %>% 
    group_by(stock_id) %>%                                           # Grouping
    mutate(R1M_Usd = lag(R1M_Usd)) %>%                              # Lag returns
    ungroup() %>%
    na.omit() %>%                                                   # Remove missing points
    pivot_wider(names_from = "stock_id", values_from = "R1M_Usd")

models <- lapply(paste0("`", stock_ids, 
                        '` ~  MKT_RF + HML + LMH'),           # Model spec
                 function(f){ lm(as.formula(f), data = data_FM,           # Call lm(.)
                                 na.action="na.exclude") %>%       
                         summary() %>%                                    # Gather the output
                         "$"(coef) %>%                                    # Keep only coefs
                         data.frame() %>%                                 # Convert to dataframe
                         dplyr::select(Estimate)}                         # Keep the estimates
                 )
betas <- matrix(unlist(models), ncol = nb_factors + 1, byrow = T) %>%     # Extract the betas
    data.frame(row.names = stock_ids)                               # Format: row names
colnames(betas) <- c("Constant", "MKT_RF", "HML", "LMH")    # Format: col names
```


```{r}
loadings <- betas %>%                            # Start from loadings (betas)
    dplyr::select(-Constant) %>%                 # Remove constant
    data.frame()                                 # Convert to dataframe             
ret <- returns %>%                               # Start from returns
    dplyr::select(-date) %>%                     # Keep the returns only
    data.frame(row.names = returns$date) %>%     # Set row names
    t()                                          # Transpose
FM_data <- cbind(loadings, ret)                  # Aggregate both
```


Then we do the second round of regression
```{r}
models <- lapply(paste("`", returns$date, "`", ' ~  MKT_RF + HML + LMH', sep = ""),
function(f){ lm(as.formula(f), data = FM_data) %>%                        # Call lm(.)
                         summary() %>%                                    # Gather the output
                         "$"(coef) %>%                                    # Keep only the coefs
                         data.frame() %>%                                 # Convert to dataframe
                         dplyr::select(Estimate)}                         # Keep only estimates
                 )
gammas <- matrix(unlist(models), ncol = nb_factors + 1, byrow = T) %>%    # Switch to dataframe
    data.frame(row.names = returns$date)                                  # & set row names
colnames(gammas) <- c("Constant", "MKT_RF", "HML", "LMH")   # Set col names
```

Visually the gamma
```{r}
gammas[2:nrow(gammas),] %>%                                         # Take gammas:
    # The first row is omitted because the first row of returns is undefined
    dplyr::select(MKT_RF, HML, LMH) %>%                             # Select 3 factors
    bind_cols(date = data_FM$date) %>%                              # Add date
    gather(key = factor, value = gamma, -date) %>%                  # Put in tidy shape
    ggplot(aes(x = date, y = gamma, color = factor)) +              # Plot
    geom_line() + facet_grid( factor~. ) +                          # Lines & facets
    scale_color_manual(values=c("#F87E1F", "#0570EA", "#F81F40")) #+ # Colors
    #coord_fixed(980) + theme_light()                                # Fix x/y ratio
```

Calculate the statistics
```{r}

library(knitr)
# Gather the main statistics
stats <-
  gammas %>%
  summarise(mean(MKT_RF), mean(HML), mean(LMH), sd(MKT_RF), sd(HML), sd(LMH)) %>%
  round(3)

stats

# Calculate t-stats
r <- t.test(gammas$MKT_RF)
r$statistic
r <- t.test(gammas$HML)
r$statistic
r <- t.test(gammas$LMH)
r$statistic

# Calculate coefficients
cor(gammas$MKT_RF, gammas$HML)
cor(gammas$MKT_RF, gammas$LMH)
cor(gammas$HML, gammas$LMH)
```


